Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-09 paper/main.tex

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative DAG construction method that
combines domain knowledge with data-driven insights. Our approach iteratively
identifies pair of variables with a missing edge in the DAG, and asks for the
orientation of the edge from an expert. We also propose a method to rank these
potential new edges based on their impact on the model fit. Empirical results
show that this iterative model construction approach achieves performance
comparable to automated algorithms if we are able to correctly decide the
orientation of new edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model (LLM) can
be used to determine edge orientations. We provide the implementation of our
approach in both a web tool at: <redacted for review> and a Python package
<redacted for review>.

% Our approach combines implied Conditional Independence (CI) testing with a
% measure of conditional association between variables to rank violations
% of implied CIs in a given DAG. This ranking helps prioritize modifications to
% the DAG to improve its fit to the data while utilizing expert knowledge to
% decide the orientation of new edges.

>>>>

	Causal discovery has received significant attention in the Directed
	Acyclic Graphs (DAGs) literature, leading to the development of
	numerous automated algorithms for learning DAGs from data. However,
	their adoption in applied domains remain limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to make obvious errors and output
	Markov Equivalence Classes (MECs) rather than a DAG. To assist
	researchers in constructing DAGs manually, we propose an iterative DAG
	construction approach that combines domain knowledge with data-driven
	insights. Our method leverages Conditional Independence (CI) testing to
	iteratively identify variable pairs where an edge is either missing or
	superfluous. Based on this information, we can choose to add missing
	edges with appropriate orientation based on domain knowledge or remove
	unnecessary ones. We also give a method to rank potential new edges
	based on their impact on the overall model fit. Empirical results show
	that this iterative approach achieves performance comparable to
	automated algorithms if the orientation of new edges is correctly
	determined in at least one out of two cases. Additionally, we show
	that, in the absence of an expert, a Large Language Model (LLM) can be
	used to determine edge orientations. We provide an implementation of
	our approach as both a webtool: <redacted for review> and a Python
	package <redacted for review>.
